---
title: "Curriculum Vitae"
about:
  template: jolla
---

```{r setup, include=FALSE, message=F, warning=F}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(tidyverse)
```

## Shane A. Scaggs

<p style="text-align: left;">

Presidential Fellow<br> [Human Complexity Lab](https://u.osu.edu/hclab/)<br> Department of Anthropology<br> The Ohio State University<br> scaggs.32\@osu.edu

</p>

<br>

### Research Interests

Cooperative Networks $\times$ Landscape Mosaics $\times$ Food Webs $\times$ Ecological Anthropology $\times$ Complex Systems $\times$ Community and Landscape Ecology $\times$ Computational Social Science $\times$ Network Science

<br>

### Education

```{r,echo=F}
edu <- read.csv("C:/Users/scaggs.32/OneDrive - The Ohio State University/Professional/sascaggs.github.io/csv/edu.csv")
kbl(edu, format = 'html', align = 'l') %>% 
  row_spec(nrow(edu), extra_latex_after = "\\arrayrulecolor{white}") %>%
  row_spec(0, bold = T) %>%
  column_spec(1, width = '0.4in') %>%
  column_spec(2, width = '2in') %>%
  column_spec(3, width = '2in') %>%
  column_spec(4, width = '1.5in')
```

### Publications

```{r,echo=F}
library(kableExtra)

bold_text <- function(text, to_bold) {
  gsub(
    to_bold, 
    paste0("\\\\textbf{", to_bold, "}"), 
    text, 
    fixed = TRUE
  )
}
pubs <- openxlsx::read.xlsx(xlsxFile = "C:/Users/scaggs.32/OneDrive - The Ohio State University/Professional/sascaggs.github.io/csv/pubs.xlsx", sheet=3)
pubs$Citation = paste0(pubs$Authors, ' ', 
                       pubs$Title, ' ',
                       '\\textit{', paste0(pubs$Journal),'} ', 
                       pubs$Volume, ' ', 
                       pubs$DOI)
pubs$Citation = gsub('_', '\\\\_', pubs$Citation)
pubs$Citation = gsub('Scaggs, S. A.', '\\\\textbf{Scaggs, S. A.}', pubs$Citation)

#pubs$Citation = gsub('/', '\\\\/', pubs$Citation)
#pubs$Citation = gsub('[', '\\\\[', pubs$Citation)
#pubs$Citation = gsub(']', '\\\\]', pubs$Citation)

pubs = pubs |> arrange(desc(Year))
kbl(pubs[,c('Year','Citation')], format = 'html', align = 'l', linesep = "\\addlinespace", longtable = T, row.names = NA, escape=F) %>% 
  kable_styling(latex_options = c('HOLD_position','repeat_header')) %>%
  row_spec(nrow(pubs), extra_latex_after = "\\arrayrulecolor{white}") %>%
  row_spec(0, bold = T) %>%
  column_spec(1, width = '1in') %>%
  column_spec(2, width = '5in')


```

## Contributions

```{r, echo=F, message=F, warning=F, fig.width=8, fig.height=4, fig.cap="Contribution matrix.", fig.align='center'}
cont <- openxlsx::read.xlsx(xlsxFile = "C:/Users/scaggs.32/OneDrive - The Ohio State University/Professional/sascaggs.github.io/csv/pubs.xlsx", sheet=2)

cont %>% 
  gather(key=key, value=value, -Contribution) %>%
  mutate(value = ifelse(is.na(value)==T, 'No', 'Yes')) %>%
  ggplot() + geom_tile(aes(x=key, y=Contribution, fill=factor(value),), color='white',size=1) + 
  scale_fill_manual(values = c('white','#09c9c988')) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 40, hjust = 1), 
        axis.text = element_text(color='black'),
        text = element_text(size=14, color='black'), 
        panel.grid = element_blank()) + 
  labs(x=NULL, y=NULL, fill='Contributed?')  
```

### Awards, Fellowships, and Funding Support

```{r,echo=F}
funds <- read.csv("C:/Users/scaggs.32/OneDrive - The Ohio State University/Professional/sascaggs.github.io/csv/funds.csv")
#colnames(funds) <- c('Year','Title','Agency','Amount')
funds <- funds[, 1:2]

kbl(funds, format = 'html', align = 'l', linesep = "\\addlinespace", longtable = T) %>% 
  kable_styling(latex_options = c('HOLD_position','repeat_header')) %>%
  row_spec(nrow(funds), extra_latex_after = "\\arrayrulecolor{white}") %>%
  row_spec(0, bold = T) %>%
  column_spec(1, width = '1in') %>%
  column_spec(2, width = '5in')
```

### Conference Proceedings

```{r,echo=F}
conf <- read.csv("C:/Users/scaggs.32/OneDrive - The Ohio State University/Professional/sascaggs.github.io/csv/conf.csv")
#colnames(conf) <- c('Year','Title','Conference','Type')
conf <- conf %>% arrange(desc(Year))


kbl(conf[,1:2], format = 'html', align = 'l', linesep = "\\addlinespace", longtable = T) %>% 
  kable_styling(latex_options = c('HOLD_position','repeat_header')) %>%
  row_spec(nrow(conf), extra_latex_after = "\\arrayrulecolor{white}") %>%
  row_spec(0, bold = T) %>%
  column_spec(1, width = '1in') %>%
  column_spec(2, width = '5in') 
```

### Teaching Experience

```{r, echo=FALSE}
teach <- read.csv("C:/Users/scaggs.32/OneDrive - The Ohio State University/Professional/sascaggs.github.io/csv/teach.csv")
teach$Term = paste0(teach$Term, ' ', teach$Year)
teach <- teach[ , c(2,3,4,5)]


kable(teach, format = 'html', align = 'l', linesep = "\\addlinespace", longtable = T) %>% 
  kable_styling(latex_options = c('HOLD_position','repeat_header')) %>%
  row_spec(nrow(teach), extra_latex_after = "\\arrayrulecolor{white}") %>%
  row_spec(0, bold = T) %>%
  column_spec(1, width = '1in') %>%
  column_spec(2, width = '2.25in') %>%
  column_spec(3, width = '2.25in') %>%
  column_spec(4, width = '0.5in') 
```

### Service and Outreach

```{r, echo=F}
serv <- read.csv("C:/Users/scaggs.32/OneDrive - The Ohio State University/Professional/sascaggs.github.io/csv/serv.csv")
serv <- serv %>% select(Year, Role)

kbl(serv, format = 'latex', align = 'l', linesep = "\\addlinespace", longtable = T) %>% 
  kable_styling(latex_options = c('HOLD_position','repeat_header')) %>%
  row_spec(nrow(serv), extra_latex_after = "\\arrayrulecolor{white}") %>%
  row_spec(0, bold = T) %>%
  column_spec(1, width = '1.25in') %>%
  column_spec(2, width = '4.75in')

```

