---
title: "Curriculum Vitae"
about:
  template: jolla
---

```{r setup, include=FALSE, message=F, warning=F}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(tidyverse)
```

## Shane A. Scaggs

<p style="text-align: left;">

Presidential Fellow<br> [Human Complexity Lab](https://u.osu.edu/hclab/)<br> Department of Anthropology<br> The Ohio State University<br> scaggs.32\@osu.edu

</p>

<br>

### Research Interests

Human Subsistence $\times$ Cooperation $\times$ Community and Landscape Ecology $\times$ Computational Social Science $\times$ Complex Systems $\times$ Network Science

<br>

### Education

```{r,echo=F}
edu <- read.csv(file = "C:/Users/scaggs.32/OneDrive - The Ohio State University/Profession Development/sascaggs.github.io/csv/edu.csv")
edu %>% kbl(align = 'l') %>% column_spec(1, width = '0.5in')
```

### Publications

```{r,echo=F}
pubs <- openxlsx::read.xlsx("C:/Users/scaggs.32/OneDrive - The Ohio State University/Profession Development/sascaggs.github.io/csv/pubs.xlsx", sheet=1)
colnames(pubs) <- c('Year','Citation','DOI','Type')

pubs$Citation <- paste0(pubs$Citation, ' ', pubs$DOI)

pubs[,1:2] %>% 
  mutate(Citation = cell_spec(Citation, "html", link = pubs$DOI)) |> 
  kable(align = 'l', "html", escape = F) %>%
  column_spec(1, width = '0.5in')

```

## Contributions

```{r, echo=F, message=F, warning=F, fig.width=8, fig.height=4, fig.cap="Contribution matrix.", fig.align='center'}
cont <- openxlsx::read.xlsx("C:/Users/scaggs.32/OneDrive - The Ohio State University/Profession Development/sascaggs.github.io/csv/pubs.xlsx", sheet=2)

cont %>% 
  gather(key=key, value=value, -Contribution) %>%
  mutate(value = ifelse(is.na(value)==T, 'No', 'Yes')) %>%
  ggplot() + geom_tile(aes(x=key, y=Contribution, fill=factor(value)), color='black',size=1) + 
  scale_fill_manual(values = c('white','magenta')) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 40, hjust = 1), 
        axis.text = element_text(color='black'),
        text = element_text(size=14, color='black')) + 
  labs(x=NULL, y=NULL, fill='Contributed?')  
```

### Funding Support

```{r,echo=F}
funds <- read.csv(file = "C:/Users/scaggs.32/OneDrive - The Ohio State University/Profession Development/sascaggs.github.io/csv/funds.csv")
colnames(funds) <- c("Year","Title","Funding Agency","Amount")
funds %>% kbl(align = 'l') %>% column_spec(1, width = '0.5in')
```

### Conference Proceedings

```{r,echo=F}
conf <- read.csv(file = "C:/Users/scaggs.32/OneDrive - The Ohio State University/Profession Development/sascaggs.github.io/csv/conf.csv")
colnames(conf) <- c("Year","Title","Meeting","Type")
conf[,1:4] %>% 
  arrange(desc( Year)) %>%
  kbl(align = 'l') %>% 
  column_spec(1, width = '0.5in')

```
